\chapter{Compiler Structure}
\label{chapter:compiler structure}

Most Compilers that are used nowadays are built of three basic components which handle different steps in the process of converting human-readable programming language to machine-readable machine code. As does the GNU Compiler collection (GCC) which also can be seen as made of three main parts. The so called front-end, middle-end and back-end.
All three parts work more or less independently from each other and communicate over a compiler-specific „language“, which is described a the Intermediate Representation (IR). It is typically never seen by the user and exists for a fact in many different forms [reference] one of which is Register Transfer Language (RTL), which is the lowest-level IR used by GCC. It is the most interesting IR when working on a back-end and will get more attention later in this report. But in before that we need to understand the structure an functioning of a compiler in general.

The first mentioned Front-end resembles the main interaction point between the human programmer and the machine. Front-ends are usually divided by their respective programming languages such as C, C++, Java…  and have the main task of converting any programming language into unified IR, that can be passed to the middle-end in GIMPLE or GENERIC language. Therefore no matter which language you prefer, in an ideal case code, that is syntactically identical, should not differ after it is processed by the front-end. This is due to the goal of compilers such as GCC and LLVM to support as many languages and machines as reasonably possible while offering the equally good optimization and saving themselves overall work. It obvious that a single compiler for every combination of language and machine would simply not be practical, especially as the optimization taking place in the middle-end follows the same rules for pretty much any architecture.
The middle-end main task is basically this sort of optimization, and makes for the main difference between compilers as most compilers offer the same range of front-ends and back-ends. (Part about the optimizations taking place in the middle-end). After all optimizations are through, the middle-end passes IR in form of RTL to the back-end. As you can see the middle-end rarely needs to be modified except for fundamental changes in the compilers architecture such as new kinds of optimizations and „multiple memory handling“ (also Harvard-Architecture (vielleicht))
The back-end is responsible for the final steps of the compilation process as it translates the general RTL IR into specific Assembly commands. It uses some sort of table of available assembly instructions, that is provided, and finds the best fitting instructions. GCC for example uses a Lisp-like language (is this RTL?) that uses something called insns. These combine different properties with the  Assembly commands like an equivalent set of actions that are executed, the  operands and the constraints it must satisfy. These will be further explained later. The back-end also contains the the code which implements processor-specific built-in functions.
Depending on the Compiler architecture the back-end it finally emits IR to the assembler which emits the machine code in assembly or emits assembly itself. Then finally the linker links the assembly code of all program files together and substitutes the offset addresses with absolut addresses to generate the final machine code.
reload
register spilling
register handling
endianess
wordsize

For the rest of this report we will deal more specifically with the combination of the C front-end and the RS/6000 back-end which is also used for the Power ISA which itself is the basis for the PPU. As exemplary built-in functions we will use the AltiVec vector extension for the rs/6000 architecture because the PPU has its own vector extension as well and ultimately this is meant to help extend the set of vector functions for the PPU.
